Approximating the matrix exponential derivative with respect to the CTMC generator matrix, with application to global SARS-CoV-2

### 2023-03-09 manuscript/main.tex
need to start manuscript

\title{Approximating the matrix exponential derivative with respect to a structured matrix}

@@

\author{Andrew J.~Holbrook}

\affil{UCLA Biostatistics}

@@

Let $\QQ=[q_{ij}]$ be the rate matrix for an $S$-dimensional continuous time Markov chain, where $q_{ij}>0$ for $i\neq j$, and $q_{ii} = -\sum_{j\neq i} q_{ij}$.  In our application of interest, $\QQ$ is almost surely diagonalizable, i.e., $\QQ=\MM \DD \MM^{-1}$ for 
\begin{align*}
	\DD = \begin{pmatrix}
		d_1& 0 &\dots & & \\
		0  & d_2 &0  & \dots& \\
		\vdots &0 & \ddots & \\
		& \vdots& & d_{S-1} &0 \\
		&&&0&0
	\end{pmatrix} \, ,
\end{align*}
where the structure of $\QQ$ implies eigenvalues
\begin{enumerate}
	\item $d_s\in \mathbb{C}$ with $\mbox{Re}(d_s)<0$ for $s\in\{1,\dots,S-1\}$, and
	\item $d_S =0$ corresponding to the eigenvector $\boldsymbol{1}=(1,\dots,1)^T$. 
\end{enumerate} 
We may further define a generalized inverse $\QQ^+=\MM \DD^{+} \MM^{-1}$ of $\QQ$ using $\DD^+$, the Moore-Penrose pseudoinverse obtained by inverting the non-zero elements of $\DD$.   Finally, we note that diagonalizability implies $\DD$ and $\DD^+$ commute, i.e., $\DD^+\DD=\DD\DD^+$.

 We are interested in efficiently approximating the directional derivative of the matrix exponential
\begin{align*}
	e^{t\QQ} = \sum_{k=0}^\infty \frac{t^k\QQ^k}{k!}
\end{align*}
for arbitrary $t>0$ using the formula \citep{najfeld1995derivatives}
\begin{align}\label{eq:deriv}
	\nabla_{\JJ} e^{t\QQ}  =  e^{t\QQ}  \sum_{k=0}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} \, .
\end{align}
Here, $\JJ$ gives the direction and $\{\JJ,\QQ^k\}$ is the matrix commutator series satisfying the recursion
\begin{align*}
	\{\JJ,\QQ^0\} &= \JJ \, , \\ 
	\{\JJ,\QQ^k\} &= [\{\JJ,\QQ^{k-1}\} , \QQ]  = \{\JJ,\QQ^{k-1}\} \QQ - \QQ \{\JJ,\QQ^{k-1}\} \, .
\end{align*}
Thus, the first-order approximation to derivative \eqref{eq:deriv} is 
\begin{align}\label{eq:firstOrder}
	\widetilde{\nabla}_{\JJ} e^{t\QQ}= t e^{t\QQ}\JJ \, .
\end{align}
We want to bound the truncation error
\begin{align*}
\mbox{err}(t,\QQ,\JJ)	:=\lVert \nabla_{\JJ} e^{t\QQ} - \widetilde{\nabla}_{\JJ} e^{t\QQ} \rVert \, ,
\end{align*}
 where $\lVert \cdot \rVert$ denotes an appropriate matrix norm.  In this context, we find that naive bounds (Appendix \ref{sec:loose}) that do not exploit the structure of $\QQ$  are rather loose (Figure \ref{fig:frob}).  
 In our efforts to fully leverage the structure of $\QQ$, we will make use of the general formula for the matrix commutator series  \citep{volkin1968iterated}
\begin{align}\label{eq:commSeries}
	\{ \JJ, \QQ^k\} = \sum_{r=0}^k (-1)^{k+r}  {k \choose r} \QQ^{k-r} \JJ \QQ^r  \, .
\end{align}

\subsection{An empirically tight asymptotic estimate}

We begin by noting
\begin{align}\label{eq:ub1}
	\mbox{err}(t,\QQ,\JJ) &=  \lVert  e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} \rVert \\ \nonumber
	&\leq  \lVert  e^{t\QQ}\JJ  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \QQ^k \rVert +   \lVert  e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} -\JJ\QQ^k\rVert \\ \nonumber
		&=  \lVert  e^{t\QQ}\JJ \QQ^+  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \QQ^{k+1} \rVert +   \lVert  e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} -\JJ\QQ^k\rVert \\ \nonumber
		&=  \lVert  e^{t\QQ}\JJ \QQ^+ (e^{t\QQ} - t\QQ -\II) \rVert +   \lVert  e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} -\JJ\QQ^k\rVert \\ \nonumber
			&\leq \lVert  e^{t\QQ}\JJ \QQ^+ ( t\QQ +\II) \rVert  + \lVert  e^{t\QQ}\JJ \QQ^+ e^{t\QQ} \rVert  +   \lVert  e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} -\JJ\QQ^k\rVert \\ \nonumber
			&=: \mathfrak{a} + \mathfrak{b} + \mathfrak{c} \, .
\end{align}
Empirical experiments using the Frobenius and infinity norms (Figures \ref{fig:frob} and \ref{fig:inf}, respectively) show that the first term
\begin{align}
	 \mathfrak{a} = \left\Vert e^{t\QQ} \JJ \QQ^+(t\QQ+\II)  \right\Vert \, ,
\end{align}
although not a strict upper bound, is asymptotically tight for the true error as $t\rightarrow \infty$. We would, therefore, like to show that the terms $\mathfrak{b}$ and $\mathfrak{c}$ go to zero as $t$ diverges.

\section{Using the structure of $\QQ$}

We use the fact that
\begin{align*}
	e^{t\QQ}\QQ = \MM \exp(t\DD) \DD \MM^{-1} =  \MM \begin{pmatrix}
	e^{td_1}	d_1& 0 &\dots & & \\
		0  & e^{td_2}d_2 &0  & \dots& \\
		\vdots &0 & \ddots & \\
		& \vdots& & e^{td_{S-1}}d_{S-1} &0 \\
		&&&0&0
	\end{pmatrix} \MM^{-1} \, .
\end{align*}
Thus, assuming use of a submultiplicative norm leads to the result
\begin{align*}
	\lVert e^{t\QQ}\QQ \rVert &\leq \left\Vert  \begin{pmatrix}
		e^{td_1}	d_1& 0 &\dots & & \\
		0  & e^{td_2}d_2 &0  & \dots& \\
		\vdots &0 & \ddots & \\
		& \vdots& & e^{td_{S-1}}d_{S-1} &0 \\
		&&&0&0
	\end{pmatrix}  \right\Vert  \lVert \MM\rVert   \lVert \MM^{-1}\rVert  \, .
\end{align*}
For, e.g., the Frobenius norm, we have
\begin{align*}
\lVert e^{t\QQ}\QQ \rVert_F	\leq \sqrt{S-1} \max_s \{|e^{td_{s}} d_{s}| \}  \lVert \MM\rVert_F   \lVert \MM^{-1}\rVert_F  \, ,
\end{align*}
where we  use the fact that $\lVert \cdot \rVert_F \leq \sqrt{r} \lVert \cdot \rVert_2$ for any rank $r$ matrix and $\lVert \cdot \rVert_2$ the spectral norm.  We therefore write
\begin{align}
	\lVert e^{t\QQ}\QQ \rVert_F = \order{\max_s \left\{|e^{td_{s}} d_{s}| \right\}} \,,  \quad 	\lVert e^{t\QQ}\QQ^+ \rVert_F = \order{\max_s \left\{ \left| e^{td_{s}} \frac{1}{d_{s}}\right| \right\}} \, .
\end{align}
Because $\mbox{Re}(d_s)<0$ for $s\in \{1,\dots,S-1\}$, we have
\begin{align}
	\lim_{t\rightarrow \infty}  	\lVert e^{t\QQ}\QQ \rVert_F = 	\lim_{t\rightarrow \infty}  	\lVert e^{t\QQ}\QQ^+ \rVert_F = 0\, ,
\end{align}
and, by equivalence of norms, we have
\begin{align}
	\lim_{t\rightarrow \infty}  	\lVert e^{t\QQ}\QQ \rVert = 	\lim_{t\rightarrow \infty}  	\lVert e^{t\QQ}\QQ^+ \rVert = 0\, ,
\end{align}
in general for any matrix norm $\lVert \cdot \rVert$. Moreover, the matrix $e^{t\QQ}$ is a stochastic matrix with rows summing to unity. Therefore $\lVert e^{t\QQ}\rVert_\infty = 1$, and, by norm equivalence, $\lVert e^{t\QQ}\rVert = \mathcal{O}(1)$.
 These results lead to the following fact for term $\mathfrak{b}$ in the upper bound of \eqref{eq:ub1}:
\begin{align*}
	\lim_{t\rightarrow \infty} \lVert  e^{t\QQ}\JJ \QQ^+ e^{t\QQ} \rVert \leq  \lim_{t\rightarrow \infty} \lVert  e^{t\QQ} \rVert \lVert \JJ  \rVert \lVert \QQ^+ e^{t\QQ} \rVert = 0 \, .
\end{align*}
Thus, we obtain the intermediary result for $t$ large:
\begin{align*}
\lim_{t\rightarrow \infty}\mbox{err}(t,\QQ,\JJ) \leq \lim_{t\rightarrow \infty}  \mathfrak{a}+ \mathfrak{c} \, .
\end{align*}
It remains to show that $\mathfrak{c}$ goes to zero as $t$ diverges, i.e., 
\begin{align}
	\lim_{t\rightarrow \infty} \lVert  e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} -\JJ\QQ^k\rVert = 0\, .
\end{align}

\section{Random thoughts}

\subsection{Another form}

\newcommand{\PPsi}{\boldsymbol{\Psi}}

\citet{najfeld1995derivatives} use the integral representation to prove the interesting result
\begin{align*}
	\nabla_{\JJ} e^{t\QQ} = \MM \left(  \left(\MM^{-1} \JJ \MM\right) \circ \PPsi(t)  \right) \MM^{-1} \, ,
\end{align*}
where $\PPsi(t)$ has elements
\begin{align*}
		[\PPsi(t)]_{ij} = \bigg\{ \begin{array}{cc}
		t e^{td_i}\, ,	&  d_i = d_j \\
		\frac{e^{td_i}-e^{td_j}}{d_i-d_j} \, ,& d_i \neq d_j 
	\end{array} \, .
\end{align*}
For our application, the eigenvalues equal one another with measure zero.  Taking $t$ large, we have
\begin{align*}
	[\PPsi(t)]_{ij} = \bigg\{ \begin{array}{cc}
		t e^{td_i}\, ,	&  i = j \\
		0 \, ,& i \neq j 
	\end{array} \, .
\end{align*}
It follows that, again for $t$ large,
\begin{align*}
	\mbox{err}(t,\QQ,\JJ) &\approx \lVert   \MM \left(  \left(\MM^{-1} \JJ \MM\right) \circ t e^{t\DD}  \right) \MM^{-1}       -  t e^{t\QQ} \JJ            \rVert \\
	&= t\lVert   \MM \rVert \lVert \left(  \left(\MM^{-1} \JJ \MM\right) \circ  e^{t\DD}  \right) \MM^{-1}       -   e^{t\DD}\MM^{-1} \JJ            \rVert \, .
\end{align*}

\appendix

\section{Loose bounds}\label{sec:loose}

The following loose bounds do not leverage the structure of the CTMC generator matrix $\QQ$.

\subsection{Upper bound 1}

For a submultiplicative matrix norm, we have
\begin{align*}
	\Vert \{\MM_1, \MM_2^k \}  \Vert \leq 2^k \Vert \MM_1\Vert \Vert\MM_2 \Vert ^k \, .
\end{align*}
We use this to obtain our first upper bound on $err_{ij}(\QQ)$:
\begin{align*}
	err_{ij}(\QQ) &= \left\Vert e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} \right\Vert  \\
	&\leq \Vert e^{t\QQ} \Vert \sum_{k=1}^\infty \left\Vert \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} \right\Vert  \\
	&=  \Vert e^{t\QQ} \Vert \sum_{k=1}^\infty \frac{t}{k+1} \frac{(2t \Vert \QQ \Vert)^k}{k!}\\
	&= \Vert e^{t\QQ} \Vert   \frac{\exp(2t\Vert \QQ \Vert)-2t\Vert\QQ\Vert-1}{2\Vert \QQ \Vert} 
\end{align*}

\subsection{Upper bound 2}

We start with another representation of the derivative \citep{al2009computing}:
\begin{align*}
	\nabla_{\JJ} e^{t\QQ} = \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^k \QQ^{p-1} \JJ \QQ^{k-p} 
\end{align*}
and the fact that
the fact that, for the Frobenius norm $\Vert \cdot \Vert_F$, we have \citep{wu2010short}
\begin{align*}
	\Vert [\MM_1,\MM_2] \Vert_F \leq \sqrt{2} \Vert \MM_1 \Vert_F \Vert \MM_2\Vert_F \, .
\end{align*}
Using these results, the truncation error may be rewritten
\begin{align*}
	err_{ij}(\QQ)  &=\left\Vert - te^{t\QQ} \JJ  +  \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^k \QQ^{p-1} \JJ \QQ^{k-p}\right\Vert_F \\
	&= \left\Vert \sum_{k=1}^\infty \frac{t^k}{k!} \left( \left( \sum_{p=1}^k \QQ^{p-1} \JJ \QQ^{k-p} \right) - k \QQ^{k-1} \JJ \right)  \right\Vert_F\\
	&= \left\Vert \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^k \QQ^{p-1} \JJ \QQ^{k-p}  -  \QQ^{k-1} \JJ \right\Vert_F \\
	&=\left\Vert  \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^{k-1} \QQ^{p-1} \JJ \QQ^{k-p}  -  \QQ^{k-1} \JJ\right\Vert_F \\
	&=\left\Vert  \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^{k-1} \QQ^{p-1}\left( \JJ \QQ^{k-p}  -  \QQ^{k-p} \JJ \right)\right\Vert_F \\
	&= \left\Vert  \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^{k-1} \QQ^{p-1}[\JJ, \QQ^{k-p}] \right\Vert_F  \\
	&\leq  \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^{k-1} \Vert \QQ^{p-1} \Vert_F \Vert[\JJ, \QQ^{k-p}]\Vert_F 
	\leq \sqrt{2}  \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^{k-1} \Vert \QQ^{k-1} \Vert_F  \\
	&= \sqrt{2}  \sum_{k=1}^\infty \frac{t^k}{k!} (k-1) \Vert \QQ \Vert_F^{k-1} = \sqrt{2} \frac{e^{t\Vert \QQ\Vert_F}(t\Vert\QQ\Vert_F - 1) + 1}{\Vert \QQ \Vert_F} \, .
\end{align*}

\begin{figure}[!p]
	\centering
	\includegraphics[width=\linewidth]{frobeniusScaling.pdf}
	\caption{Anecdotal (single simulation) error profiles with change in time $t$ and dimension. Errors calculated using Frobenius norm.}\label{fig:frob}
\end{figure}

\begin{figure}[!p]
	\centering
	\includegraphics[width=\linewidth]{infinityScaling.pdf}
	\caption{Anecdotal (single simulation) error profiles with change in time $t$ and dimension. Errors calculated using infinity norm.}\label{fig:inf}
\end{figure}

>>>>

\title{Approximating the matrix exponential derivative with respect to the CTMC generator matrix, with application to global SARS-CoV-2}

@@

\author[1]{Andrew J.~Holbrook}
\author[1]{Andrew F.~Magee}
\author[2]{Nathan E.~Glatt-Holtz}
\author[1,3,4]{Marc A.~Suchard}

 \affil[1]{Department of Biostatistics, University of California, Los Angeles}
 \affil[2]{Department of Mathematics, Tulane University}
 \affil[3]{Department of Biomathematics, University of California, Los Angeles}
 \affil[4]{Department of Human Genetics, University of California, Los Angeles}

@@

\section{Global spread of SARS-CoV-2}

%Let $\QQ=[q_{ij}]$ be the rate matrix for an $S$-dimensional continuous time Markov chain, where $q_{ij}>0$ for $i\neq j$, and $q_{ii} = -\sum_{j\neq i} q_{ij}$.  In our application of interest, $\QQ$ is almost surely diagonalizable, i.e., $\QQ=\MM \DD \MM^{-1}$ for 
%\begin{align*}
%	\DD = \begin{pmatrix}
%		d_1& 0 &\dots & & \\
%		0  & d_2 &0  & \dots& \\
%		\vdots &0 & \ddots & \\
%		& \vdots& & d_{S-1} &0 \\
%		&&&0&0
%	\end{pmatrix} \, ,
%\end{align*}
%where the structure of $\QQ$ implies eigenvalues
%\begin{enumerate}
%	\item $d_s\in \mathbb{C}$ with $\mbox{Re}(d_s)<0$ for $s\in\{1,\dots,S-1\}$, and
%	\item $d_S =0$ corresponding to the eigenvector $\boldsymbol{1}=(1,\dots,1)^T$. 
%\end{enumerate} 
%We may further define a generalized inverse $\QQ^+=\MM \DD^{+} \MM^{-1}$ of $\QQ$ using $\DD^+$, the Moore-Penrose pseudoinverse obtained by inverting the non-zero elements of $\DD$.   Finally, we note that diagonalizability implies $\DD$ and $\DD^+$ commute, i.e., $\DD^+\DD=\DD\DD^+$.
%
%
% We are interested in efficiently approximating the directional derivative of the matrix exponential
%\begin{align*}
%	e^{t\QQ} = \sum_{k=0}^\infty \frac{t^k\QQ^k}{k!}
%\end{align*}
%for arbitrary $t>0$ using the formula \citep{najfeld1995derivatives}
%\begin{align}\label{eq:deriv}
%	\nabla_{\JJ} e^{t\QQ}  =  e^{t\QQ}  \sum_{k=0}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} \, .
%\end{align}
%Here, $\JJ$ gives the direction and $\{\JJ,\QQ^k\}$ is the matrix commutator series satisfying the recursion
%\begin{align*}
%	\{\JJ,\QQ^0\} &= \JJ \, , \\ 
%	\{\JJ,\QQ^k\} &= [\{\JJ,\QQ^{k-1}\} , \QQ]  = \{\JJ,\QQ^{k-1}\} \QQ - \QQ \{\JJ,\QQ^{k-1}\} \, .
%\end{align*}
%Thus, the first-order approximation to derivative \eqref{eq:deriv} is 
%\begin{align}\label{eq:firstOrder}
%	\widetilde{\nabla}_{\JJ} e^{t\QQ}= t e^{t\QQ}\JJ \, .
%\end{align}
%We want to bound the truncation error
%\begin{align*}
%\mbox{err}(t,\QQ,\JJ)	:=\lVert \nabla_{\JJ} e^{t\QQ} - \widetilde{\nabla}_{\JJ} e^{t\QQ} \rVert \, ,
%\end{align*}
% where $\lVert \cdot \rVert$ denotes an appropriate matrix norm.  In this context, we find that naive bounds (Appendix \ref{sec:loose}) that do not exploit the structure of $\QQ$  are rather loose (Figure \ref{fig:frob}).  
% In our efforts to fully leverage the structure of $\QQ$, we will make use of the general formula for the matrix commutator series  \citep{volkin1968iterated}
%\begin{align}\label{eq:commSeries}
%	\{ \JJ, \QQ^k\} = \sum_{r=0}^k (-1)^{k+r}  {k \choose r} \QQ^{k-r} \JJ \QQ^r  \, .
%\end{align}
%
%\subsection{An empirically tight asymptotic estimate}
%
%We begin by noting
%\begin{align}\label{eq:ub1}
%	\mbox{err}(t,\QQ,\JJ) &=  \lVert  e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} \rVert \\ \nonumber
%	&\leq  \lVert  e^{t\QQ}\JJ  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \QQ^k \rVert +   \lVert  e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} -\JJ\QQ^k\rVert \\ \nonumber
%		&=  \lVert  e^{t\QQ}\JJ \QQ^+  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \QQ^{k+1} \rVert +   \lVert  e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} -\JJ\QQ^k\rVert \\ \nonumber
%		&=  \lVert  e^{t\QQ}\JJ \QQ^+ (e^{t\QQ} - t\QQ -\II) \rVert +   \lVert  e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} -\JJ\QQ^k\rVert \\ \nonumber
%			&\leq \lVert  e^{t\QQ}\JJ \QQ^+ ( t\QQ +\II) \rVert  + \lVert  e^{t\QQ}\JJ \QQ^+ e^{t\QQ} \rVert  +   \lVert  e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} -\JJ\QQ^k\rVert \\ \nonumber
%			&=: \mathfrak{a} + \mathfrak{b} + \mathfrak{c} \, .
%\end{align}
%Empirical experiments using the Frobenius and infinity norms (Figures \ref{fig:frob} and \ref{fig:inf}, respectively) show that the first term
%\begin{align}
%	 \mathfrak{a} = \left\Vert e^{t\QQ} \JJ \QQ^+(t\QQ+\II)  \right\Vert \, ,
%\end{align}
%although not a strict upper bound, is asymptotically tight for the true error as $t\rightarrow \infty$. We would, therefore, like to show that the terms $\mathfrak{b}$ and $\mathfrak{c}$ go to zero as $t$ diverges.
%
%
%
%
%
%
%
%
%\section{Using the structure of $\QQ$}
%
%We use the fact that
%\begin{align*}
%	e^{t\QQ}\QQ = \MM \exp(t\DD) \DD \MM^{-1} =  \MM \begin{pmatrix}
%	e^{td_1}	d_1& 0 &\dots & & \\
%		0  & e^{td_2}d_2 &0  & \dots& \\
%		\vdots &0 & \ddots & \\
%		& \vdots& & e^{td_{S-1}}d_{S-1} &0 \\
%		&&&0&0
%	\end{pmatrix} \MM^{-1} \, .
%\end{align*}
%Thus, assuming use of a submultiplicative norm leads to the result
%\begin{align*}
%	\lVert e^{t\QQ}\QQ \rVert &\leq \left\Vert  \begin{pmatrix}
%		e^{td_1}	d_1& 0 &\dots & & \\
%		0  & e^{td_2}d_2 &0  & \dots& \\
%		\vdots &0 & \ddots & \\
%		& \vdots& & e^{td_{S-1}}d_{S-1} &0 \\
%		&&&0&0
%	\end{pmatrix}  \right\Vert  \lVert \MM\rVert   \lVert \MM^{-1}\rVert  \, .
%\end{align*}
%For, e.g., the Frobenius norm, we have
%\begin{align*}
%\lVert e^{t\QQ}\QQ \rVert_F	\leq \sqrt{S-1} \max_s \{|e^{td_{s}} d_{s}| \}  \lVert \MM\rVert_F   \lVert \MM^{-1}\rVert_F  \, ,
%\end{align*}
%where we  use the fact that $\lVert \cdot \rVert_F \leq \sqrt{r} \lVert \cdot \rVert_2$ for any rank $r$ matrix and $\lVert \cdot \rVert_2$ the spectral norm.  We therefore write
%\begin{align}
%	\lVert e^{t\QQ}\QQ \rVert_F = \order{\max_s \left\{|e^{td_{s}} d_{s}| \right\}} \,,  \quad 	\lVert e^{t\QQ}\QQ^+ \rVert_F = \order{\max_s \left\{ \left| e^{td_{s}} \frac{1}{d_{s}}\right| \right\}} \, .
%\end{align}
%Because $\mbox{Re}(d_s)<0$ for $s\in \{1,\dots,S-1\}$, we have
%\begin{align}
%	\lim_{t\rightarrow \infty}  	\lVert e^{t\QQ}\QQ \rVert_F = 	\lim_{t\rightarrow \infty}  	\lVert e^{t\QQ}\QQ^+ \rVert_F = 0\, ,
%\end{align}
%and, by equivalence of norms, we have
%\begin{align}
%	\lim_{t\rightarrow \infty}  	\lVert e^{t\QQ}\QQ \rVert = 	\lim_{t\rightarrow \infty}  	\lVert e^{t\QQ}\QQ^+ \rVert = 0\, ,
%\end{align}
%in general for any matrix norm $\lVert \cdot \rVert$. Moreover, the matrix $e^{t\QQ}$ is a stochastic matrix with rows summing to unity. Therefore $\lVert e^{t\QQ}\rVert_\infty = 1$, and, by norm equivalence, $\lVert e^{t\QQ}\rVert = \mathcal{O}(1)$.
% These results lead to the following fact for term $\mathfrak{b}$ in the upper bound of \eqref{eq:ub1}:
%\begin{align*}
%	\lim_{t\rightarrow \infty} \lVert  e^{t\QQ}\JJ \QQ^+ e^{t\QQ} \rVert \leq  \lim_{t\rightarrow \infty} \lVert  e^{t\QQ} \rVert \lVert \JJ  \rVert \lVert \QQ^+ e^{t\QQ} \rVert = 0 \, .
%\end{align*}
%Thus, we obtain the intermediary result for $t$ large:
%\begin{align*}
%\lim_{t\rightarrow \infty}\mbox{err}(t,\QQ,\JJ) \leq \lim_{t\rightarrow \infty}  \mathfrak{a}+ \mathfrak{c} \, .
%\end{align*}
%It remains to show that $\mathfrak{c}$ goes to zero as $t$ diverges, i.e., 
%\begin{align}
%	\lim_{t\rightarrow \infty} \lVert  e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} -\JJ\QQ^k\rVert = 0\, .
%\end{align}
%
%
%
%\section{Random thoughts}
%
%\subsection{Another form}
%
%\newcommand{\PPsi}{\boldsymbol{\Psi}}
%
%\citet{najfeld1995derivatives} use the integral representation to prove the interesting result
%\begin{align*}
%	\nabla_{\JJ} e^{t\QQ} = \MM \left(  \left(\MM^{-1} \JJ \MM\right) \circ \PPsi(t)  \right) \MM^{-1} \, ,
%\end{align*}
%where $\PPsi(t)$ has elements
%\begin{align*}
%		[\PPsi(t)]_{ij} = \bigg\{ \begin{array}{cc}
%		t e^{td_i}\, ,	&  d_i = d_j \\
%		\frac{e^{td_i}-e^{td_j}}{d_i-d_j} \, ,& d_i \neq d_j 
%	\end{array} \, .
%\end{align*}
%For our application, the eigenvalues equal one another with measure zero.  Taking $t$ large, we have
%\begin{align*}
%	[\PPsi(t)]_{ij} = \bigg\{ \begin{array}{cc}
%		t e^{td_i}\, ,	&  i = j \\
%		0 \, ,& i \neq j 
%	\end{array} \, .
%\end{align*}
%It follows that, again for $t$ large,
%\begin{align*}
%	\mbox{err}(t,\QQ,\JJ) &\approx \lVert   \MM \left(  \left(\MM^{-1} \JJ \MM\right) \circ t e^{t\DD}  \right) \MM^{-1}       -  t e^{t\QQ} \JJ            \rVert \\
%	&= t\lVert   \MM \rVert \lVert \left(  \left(\MM^{-1} \JJ \MM\right) \circ  e^{t\DD}  \right) \MM^{-1}       -   e^{t\DD}\MM^{-1} \JJ            \rVert \, .
%\end{align*}
%
%\appendix
%
%\section{Loose bounds}\label{sec:loose}
%
%The following loose bounds do not leverage the structure of the CTMC generator matrix $\QQ$.
%
%\subsection{Upper bound 1}
%
%
%
%For a submultiplicative matrix norm, we have
%\begin{align*}
%	\Vert \{\MM_1, \MM_2^k \}  \Vert \leq 2^k \Vert \MM_1\Vert \Vert\MM_2 \Vert ^k \, .
%\end{align*}
%We use this to obtain our first upper bound on $err_{ij}(\QQ)$:
%\begin{align*}
%	err_{ij}(\QQ) &= \left\Vert e^{t\QQ}  \sum_{k=1}^\infty \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} \right\Vert  \\
%	&\leq \Vert e^{t\QQ} \Vert \sum_{k=1}^\infty \left\Vert \frac{t^{k+1}}{(k+1)!} \{\JJ,\QQ^k\} \right\Vert  \\
%	&=  \Vert e^{t\QQ} \Vert \sum_{k=1}^\infty \frac{t}{k+1} \frac{(2t \Vert \QQ \Vert)^k}{k!}\\
%	&= \Vert e^{t\QQ} \Vert   \frac{\exp(2t\Vert \QQ \Vert)-2t\Vert\QQ\Vert-1}{2\Vert \QQ \Vert} 
%\end{align*}
%
%
%\subsection{Upper bound 2}
%
%We start with another representation of the derivative \citep{al2009computing}:
%\begin{align*}
%	\nabla_{\JJ} e^{t\QQ} = \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^k \QQ^{p-1} \JJ \QQ^{k-p} 
%\end{align*}
%and the fact that
%the fact that, for the Frobenius norm $\Vert \cdot \Vert_F$, we have \citep{wu2010short}
%\begin{align*}
%	\Vert [\MM_1,\MM_2] \Vert_F \leq \sqrt{2} \Vert \MM_1 \Vert_F \Vert \MM_2\Vert_F \, .
%\end{align*}
%Using these results, the truncation error may be rewritten
%\begin{align*}
%	err_{ij}(\QQ)  &=\left\Vert - te^{t\QQ} \JJ  +  \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^k \QQ^{p-1} \JJ \QQ^{k-p}\right\Vert_F \\
%	&= \left\Vert \sum_{k=1}^\infty \frac{t^k}{k!} \left( \left( \sum_{p=1}^k \QQ^{p-1} \JJ \QQ^{k-p} \right) - k \QQ^{k-1} \JJ \right)  \right\Vert_F\\
%	&= \left\Vert \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^k \QQ^{p-1} \JJ \QQ^{k-p}  -  \QQ^{k-1} \JJ \right\Vert_F \\
%	&=\left\Vert  \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^{k-1} \QQ^{p-1} \JJ \QQ^{k-p}  -  \QQ^{k-1} \JJ\right\Vert_F \\
%	&=\left\Vert  \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^{k-1} \QQ^{p-1}\left( \JJ \QQ^{k-p}  -  \QQ^{k-p} \JJ \right)\right\Vert_F \\
%	&= \left\Vert  \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^{k-1} \QQ^{p-1}[\JJ, \QQ^{k-p}] \right\Vert_F  \\
%	&\leq  \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^{k-1} \Vert \QQ^{p-1} \Vert_F \Vert[\JJ, \QQ^{k-p}]\Vert_F 
%	\leq \sqrt{2}  \sum_{k=1}^\infty \frac{t^k}{k!} \sum_{p=1}^{k-1} \Vert \QQ^{k-1} \Vert_F  \\
%	&= \sqrt{2}  \sum_{k=1}^\infty \frac{t^k}{k!} (k-1) \Vert \QQ \Vert_F^{k-1} = \sqrt{2} \frac{e^{t\Vert \QQ\Vert_F}(t\Vert\QQ\Vert_F - 1) + 1}{\Vert \QQ \Vert_F} \, .
%\end{align*}

%\begin{figure}[!p]
%	\centering
%	\includegraphics[width=\linewidth]{frobeniusScaling.pdf}
%	\caption{Anecdotal (single simulation) error profiles with change in time $t$ and dimension. Errors calculated using Frobenius norm.}\label{fig:frob}
%\end{figure}
%
%\begin{figure}[!p]
%	\centering
%	\includegraphics[width=\linewidth]{infinityScaling.pdf}
%	\caption{Anecdotal (single simulation) error profiles with change in time $t$ and dimension. Errors calculated using infinity norm.}\label{fig:inf}
%\end{figure}
